The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort Study

### 2025-04-18 products/manuscript/Manuscript.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

# Summary/Abstract

*Write a summary of your project.*

@@

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS(here("results","edatables","edasummarytable2.rds"))
knitr::kable(resulttable)
```

@@

```{r}
#| label: tbl-summarytablesurvival
#| tbl-cap: "Summary of Survival."
#| echo: FALSE
resulttable=readRDS(here("results","edatables","edasurvivalmonthssummary.rds"))
knitr::kable(resulttable)
```

@@

```{r}
#| label: tbl-summaryage
#| tbl-cap: "Summary of age."
#| echo: FALSE
resulttable=readRDS(here("results","edatables","edaagesummary.rds"))
knitr::kable(resulttable)
```

@@

```{r}
#| label: tbl-summarymetastasis
#| tbl-cap: "Summary of metastasis."
#| echo: FALSE
resulttable=readRDS(here("results","edatables","edametsatsummary.rds"))
knitr::kable(resulttable)
```

@@

```{r}
#| label: fig-mets 
#| fig-cap: "Metatstasis"
#| echo: FALSE
knitr::include_graphics(here("results","edafigures","edametsatdxdistribution.png"))
```

@@

```{r}
#| label: tbl-summaryhrmetastasis
#| tbl-cap: "Proportional hazards analysis for different metastasis."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "mets_HR.rds"))
knitr::kable(resulttable)
```

@@

```{r}
#| label: tbl-summaryhrrace
#| tbl-cap: "Proportional hazards analysis for different Race."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "race_HR.rds"))
knitr::kable(resulttable)
```

@@

```{r}
#| label: tbl-summaryhrage
#| tbl-cap: "Proportional hazards analysis for different age group."
#| echo: FALSE
resulttable=readRDS(here("results", "tables", "age_HR.rds"))
knitr::kable(resulttable)
```

@@

Variable importance analysis revealed that age group, median household income, and year of diagnosis were the most influential predictors of survival (@fig-rsf-variable-importance).
Other important variables included race and metastasis to the bone, lung, liver, and brain.
The individualized survival prediction for a representative patient showed a steady decline in survival probability over time, reflecting the nature of long-term risk in this cohort (@fig-rsf-ind-sur-pre). @fig-rsf-con-matrix. Confusion matrix showing prediction results of the Random Survival Forest model at 12 months. The model correctly classified 709 survivors and 2,182 non-survivors, with moderate misclassification of 2,376 survivors and 111 non-survivors.

@@

```{r}
#| label: fig-rsf-variable-importance
#| fig-cap: "Random Survival Forest Variable Importance"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_variable_importance.png"))
```
```{r}
#| label: fig-rsf-ind-sur-pre
#| fig-cap: "Random Survival Forest Individualized Survival Prediction"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_survival_curve_patient1.png"))
```
```{r}
#| label: fig-rsf-con-matrix
#| fig-cap: "RSF Confusion Matrix (12 Months)"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_conf_matrix_12mo.png"))
```
{{< pagebreak >}}

@@

*Summarize what you did, what you found and what it means.*

@@

*Discuss what you perceive as strengths and limitations of your analysis.*

@@

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
Many more style files for almost any journal [are available](https://www.zotero.org/styles).
You also specify the location of your bibtex reference file in the YAML.
You can call your reference file anything you like.

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(gt)
library(webshot2)
```

@@

# Abstract

**Background:** Endometrial cancer is the most common gynecological malignancy in the United States.
While most cases are diagnosed early, stage IV disease presents significant prognostic challenges, particularly when distant metastases are involved.

**Objective:** This study aimed to evaluate clinicopathological characteristics and survival outcomes among patients with stage IV endometrial cancer, focusing on differences based on metastatic site (bone, brain, liver, lung).

**Methods:** We conducted a retrospective cohort analysis using SEER data from 2010–2015.
Kaplan-Meier and Cox proportional hazards models were used to assess survival differences, and a Random Survival Forest (RSF) model was implemented to evaluate variable importance and predict survival probabilities.

**Results:** Among 5,708 patients with stage IV endometrial cancer, lung metastases were most common (29.4%), followed by liver (14.8%), bone (9.7%), and brain (2.8%).
Median survival was 12 months.
In multivariable analysis, all metastasis types were significantly associated with poorer survival, especially brain (HR = 1.72) and liver (HR = 1.66).
Older age and Black race were also linked to worse outcomes.
The RSF model achieved an AUC of 0.74, with age group, income, and metastatic site identified as the most important survival predictors.

**Conclusions:** Metastatic site significantly influences survival in stage IV endometrial cancer.
Tailored interventions and surveillance strategies based on metastasis patterns and patient demographics may help optimize outcomes.

@@

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: false
knitr::include_graphics(here::here("results", "edafigures", "summarytable.png"))
```

@@

```{r}
#| label: tbl-summarytablesurvival
#| tbl-cap: "Summary of survival months in the study population."
#| echo: false
knitr::include_graphics(here::here("results", "edafigures", "edasurvivalmonths_table.png"))
```

@@

```{r}
#| label: tbl-summaryage
#| tbl-cap: "Summary of age group distribution."
#| echo: false
knitr::include_graphics(here::here("results", "edafigures", "edaagesummary.png"))
```

@@

```{r}
#| label: tbl-summarymetastasis
#| tbl-cap: "Summary of metastasis."
#| echo: FALSE
knitr::include_graphics(here::here("results", "edafigures", "edametsatsummary.png"))
```

@@

```{r}
#| label: fig-mets 
#| fig-cap: "Metatstasis"
#| echo: FALSE
knitr::include_graphics(here::here("results", "edafigures", "edametsatdxdistribution.png"))
```

@@

```{r}
#| label: tbl-summaryhrmetastasis
#| tbl-cap: "Proportional hazards analysis for different metastasis."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "mets_HR_table.png"))
```

@@

```{r}
#| label: tbl-summaryhrrace
#| tbl-cap: "Proportional hazards analysis for different Race."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "race_HR_table.png"))
```

@@

```{r}
#| label: tbl-summaryhrage
#| tbl-cap: "Proportional hazards analysis for different age group."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "age_HR_table.png"))
```

@@

Variable importance analysis revealed that age group, median household income, and year of diagnosis were the most influential predictors of survival (@fig-rsf-variable-importance).
Other important variables included race and metastasis to the bone, lung, liver, and brain.
The individualized survival prediction for a representative patient showed a steady decline in survival probability over time, reflecting the nature of long-term risk in this cohort (@fig-rsf-ind-sur-pre).
(@fig-rsf-con-matrix).
Confusion matrix showing prediction results of the Random Survival Forest model at 12 months.
The model correctly classified 709 survivors and 2,182 non-survivors, with moderate misclassification of 2,376 survivors and 111 non-survivors.

@@

```{r}
#| label: fig-rsf-variable-importance
#| fig-cap: "Random Survival Forest Variable Importance"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_variable_importance.png"))
```

```{r}
#| label: fig-rsf-ind-sur-pre
#| fig-cap: "Random Survival Forest Individualized Survival Prediction"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_survival_curve_patient1.png"))
```

```{r}
#| label: fig-rsf-con-matrix
#| fig-cap: "RSF Confusion Matrix (12 Months)"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "rsf_conf_matrix_12mo.png"))
```

@@

In this large SEER-based analysis of patients with stage IV endometrial cancer, we found that survival outcomes varied significantly based on the site of metastasis.
Lung metastasis was the most common at diagnosis, consistent with the known hematogenous spread of endometrial tumors to the pulmonary system.
However, brain and liver metastases were associated with the poorest prognoses, showing significantly elevated hazard ratios for mortality, even after adjusting for age, race, income, and year of diagnosis.
This suggests that not all metastatic patterns carry equal prognostic weight, and involvement of certain visceral organs may reflect more aggressive tumor biology or greater resistance to conventional treatment modalities.
These findings align with previous literature highlighting the aggressive nature and poor survival associated with visceral metastases, particularly in the central nervous system and liver, where treatment options are often limited and complications more severe [@Mao2020; @el2023patterns].

Our study also highlights important demographic disparities that influence survival.
Black patients had significantly higher mortality risk compared to White patients, a pattern consistent with prior studies on racial inequities in cancer detection, treatment access, and outcomes [@Monk2022; @mccarthy2022racial].
These disparities may reflect differences in healthcare access, socioeconomic factors, and biological variations, and underscore the need for tailored interventions to address inequities in care delivery.
The impact of age was particularly striking—patients aged ≥85 years had more than 12 times the mortality risk of younger patients.
This finding supports the well-established role of age as a dominant prognostic factor in gynecologic malignancies, not only due to physiological frailty and comorbidities but also possibly due to undertreatment in older populations [@MORICE20161094; @smith2023impact].
As the population ages, understanding and addressing these age-related disparities will be increasingly important for improving outcomes in this vulnerable group.

The application of a Random Survival Forest (RSF) model in our analysis provided additional insights beyond those offered by traditional Cox regression.
By capturing complex, non-linear relationships and variable interactions, the RSF approach identified the most influential predictors of survival—including age group, income, and metastatic site.
This machine learning method yielded an AUC of 0.74, indicating good predictive performance and offering a robust complement to the conventional modeling framework.
Notably, RSF enabled individualized survival prediction, which may have potential utility in clinical decision-making and patient counseling [@ishwaran2008random; @wager2017estimation].
Integrating such tools into routine care could enhance risk stratification, support shared decision-making, and contribute to the development of precision oncology strategies.

@@

One of the study’s major strengths is the use of the SEER registry, a high-quality, population-based dataset capturing data across a wide U.S. demographic.
This enhances the generalizability of our findings and allows for robust subgroup analyses across metastasis types and sociodemographic strata.
[@howlader2021seer] Moreover, the combination of classical survival analysis and machine learning techniques provides both interpretability and predictive power.

However, our study has several limitations.
The SEER database lacks treatment details such as chemotherapy, hormonal therapy, or immunotherapy, which may influence survival outcomes.
Also, performance status and comorbidities—key clinical predictors—are not captured in SEER.
The data only reflect metastasis at diagnosis, not accounting for progression or treatment response.
Lastly, despite adjusting for income and race, residual confounding due to unmeasured variables is possible.

@@

Our findings suggest that survival in stage IV endometrial cancer is heavily influenced by the site of metastasis, with brain and liver involvement predicting worse outcomes.
The disparities across age and race underscore the need for tailored care and equitable access to treatment.
Incorporating advanced analytics such as RSF provides a deeper understanding of survival dynamics and may inform future personalized prognostic tools.
Future research should aim to integrate molecular, treatment, and longitudinal clinical data to refine prognostic models.
These findings could also guide clinicians in risk stratification, resource allocation, and decision-making, especially for high-risk subgroups.
